---
title: "EDW-PracticalMachineLearning"
author: "edwarner"
date: "Thursday, April 23, 2015"
output: html_document
---

## Practical Machine Learning Course Project

### Executive Summary

The goal of this project is to use fitness data collected from personal activiy monitoring devices to determine how the user was performing barbell lifts.  Using a random forest as my classifier with a simple 1-fold cross-validation, I was able to classify the type of barbell lift performed with 100% accuracy on the training set. Details of my model are below.

### Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

### Description of Data

Six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E).

Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes. Participants were supervised by an experienced weight lifter to make sure the execution complied to the manner they were supposed to simulate. The exercises were performed by six male participants aged between 20-28 years, with little weight lifting experience. We made sure that all participants could easily simulate the mistakes in a safe and controlled manner by using a relatively light dumbbell (1.25kg).

Read more: http://groupware.les.inf.puc-rio.br/har#ixzz3Y8nmRPuR

###  Data Processing

```{r warning=FALSE, message=FALSE, echo=FALSE}

setwd("D:/Results/R-Data/PracticalMachineLearning")

```

Load the packages need to perform the analysis

```{r warning=FALSE, message=FALSE}

library(Hmisc)
library(caret)
library(randomForest)
library(aod)
library(ggplot2)
library(doSNOW)
library(foreach)
library(e1071)

```

Load the data ignoring columns not required for the classification (username, timestamps, etc).  Also, remove all columns that have an NA value in them.

```{r warning=FALSE, message=FALSE}

set.seed(0)

data.train <- read.csv("pml-training.csv", na.strings=c("NA", "", "#DIV/0!"))

data.test <- read.csv("pml-testing.csv", na.strings=c("NA", "", "#DIV/0!"))

data.train <- data.train[colnames(data.train[colSums(is.na(data.train)) == 0])[-(1:7)]]
data.test <- data.test[colnames(data.test[colSums(is.na(data.test)) == 0])[-(1:7)]]

```

Remove all the rows with NA from the training data.

```{r warning=FALSE, message=FALSE}

data.train <- data.train[complete.cases(data.train),]

```

Split the training set into a train and validation set for cross-validation.

```{r warning=FALSE, message=FALSE}

data.validation_indexes <- createDataPartition(y=data.train$classe, p=0.70, list=FALSE )

data.train <- data.train[data.validation_indexes,]
data.validation <- data.train[-data.validation_indexes,]

rm(data.validation_indexes)

```

Setup the x and y data frames for the random forest training

```{r warning=FALSE, message=FALSE}

data.x <- data.train[-ncol(data.train)]
data.y <- data.train$classe

```

Train a random forest with 100 trees.  Use the DoSnow and foreach package to parallelize the training using 4 threads.

```{r warning=FALSE, message=FALSE}

cl<-makeCluster(4)

registerDoSNOW(cl)

rf <- foreach(ntree=rep(25, 4), .combine=combine, .packages="randomForest") %dopar% randomForest(data.x, data.y, ntree=ntree)

stopCluster(cl)
```

###  Results

Display a quick summary of the Random Forest

```{r warning=FALSE, message=FALSE}

rf

```

Calculate the predictions for the train set.

```{r warning=FALSE, message=FALSE}

data.predict_train <- predict(rf, newdata=data.train)

confusionMatrix(data.predict_train, data.train$classe)

```

As expected, the predictions on the training set are 100% accurate.

Now, calculate the predictions for the validation set.

```{r warning=FALSE, message=FALSE}

data.predict_validation <- predict(rf, newdata=data.validation)

confusionMatrix(data.predict_validation, data.validation$classe)

```

The predictions on the validation set are  100% accurate.  With an out-of-bag error rate of 0% from my validation set, I now expect my test set error to also be 0%. Therefore, on the test set, I expect to get all the predictions correct.

Calculate the predictions for the test set.

```{r warning=FALSE, message=FALSE}

data.predict_test <- predict(rf, newdata=data.test)

data.predict_test

```

Run the submission code to create the test set submission files.

```{r warning=FALSE, message=FALSE}

pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(data.predict_test)

```

###  Appendix - A Support Vector Machine Model for comparison.

```{r warning=FALSE, message=FALSE}

svm  <- svm(data.x, data.y)

```

Display a quick summary of the Support Vector Machine

```{r warning=FALSE, message=FALSE}

svm

```

Calculate the predictions on the training set.  With an accuracy of only 94.61 percent on the training set, running the SVM with default setting on this data is not quite as accurate as the random forest.

```{r warning=FALSE, message=FALSE}

data.predict_train_svm <- predict(svm, newdata=data.train[,1:52])

confusionMatrix(data.predict_train_svm, data.train$classe)

```

Calculate the predictions on the validation set.  With an accuracy of only 94.61 percent on the validation, running the SVM with default setting on this data is not quite as accurate as the random forest.

The expected out-of-bag error rate is then 5.39%.  Therefore, on the test set, I would expect approximately 1 of the 20 predictions to be incorrect.

```{r warning=FALSE, message=FALSE}

data.predict_validation_svm <- predict(svm, newdata=data.validation[,1:52])

confusionMatrix(data.predict_validation_svm, data.validation$classe)

```

Comparing the support vector machine results to the random forest results, they are both the same.  Despite the expected error rate for the support vector machine being higher than that of the random forest, it appears the SVM predicted all the test cases correctly.  With more test cases, perhaps the accuracy differences between the two models would be better reflected.


```{r warning=FALSE, message=FALSE}

data.predict_test_svm <- predict(svm, newdata=data.test[,1:52])

```

SVM

```{r warning=FALSE, message=FALSE}

#SVM
data.predict_test_svm

```

RF

```{r warning=FALSE, message=FALSE}

#RF
data.predict_test

```